---
title: 'M248: Binomial Distribution'
author: "David Thomas"
date: "05/11/2020"
output: html_document
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

\(~\)

In the unlikely event that someone reads this, please feel free to point out errors or suggest improvements. I'm so new to coding that I have probably overlooked major statistical flaws in my preoccupation with persuading my code to produce some output ... any kind of output at all!

The binomial distribution is a discrete probability distribution, which describes the outcome of a sequence of \(n\) independent Bernoulli trials in an experiment. Each trial has only two outcomes, designated either success (1) or failure (0), or vice versa. If the probability of a successful trial is \(p\), then the probability of having \(x\) successful outcomes in an experiment of \(n\) independent trials is given by:

\[f(x) = \dbinom{n}{x} p^x (1-p)^{(n-x)} \text{ where } x=1,2,\ldots,n\]

### Example 9; page 157: Book A

In an MCQ exam of 20 questions, with each question having five options, what is the probability that a student who guesses answers at random will pass the examination if the pass mark is 10. A prerequisite of Bernoulli trials is that an answer to any particular question is independent of an answer to any other question, which is the case here.

The probability of picking the correct option is 1/5 (0.2) for each question.

### Solution

The probability mass function \(p(t) = P(T = t)\) for any particular score (e.g. zero) is given by 'dbinom':

```{r}
# dbinom(x=0, size=20, prob=0.2) where x is number of successes, size is number of trials,
# and prob is the probability of success in each trial.
(score <- dbinom(x=0, size=20, prob=0.2))
```

The total number of correct answers given to the 20 questions is a random variable having a binomial
distribution with parameters \(n = 20\) and \(p = 0.2\), \(T \sim B(20, 0.2)\). If the pass mark is 10, then the probability of scoring \(<10\) (i.e. fail) is given by:

\[P(T ≤9) = P(T = 0) + P(T = 1) + P(T = 2) + \ldots + P(T = 9)\]

To find the cumulative probability of having nine or fewer correct answers by random guesses (the cumulative distribution function of \(T\), \(\; F(T)\)), which for any discrete distribution is calculated by addition, we can sequentially apply the function 'dbinom' with $x=0,1,\ldots,9$:

```{r}
(score <-  dbinom(x=0, size=20, prob=0.2) + 
          dbinom(x=1, size=20, prob=0.2) + 
          dbinom(x=2, size=20, prob=0.2) + 
          dbinom(x=3, size=20, prob=0.2) + 
          dbinom(x=4, size=20, prob=0.2) + 
          dbinom(x=5, size=20, prob=0.2) + 
          dbinom(x=6, size=20, prob=0.2) + 
          dbinom(x=7, size=20, prob=0.2) + 
          dbinom(x=8, size=20, prob=0.2) + 
          dbinom(x=9, size=20, prob=0.2))
```

Alternatively, we can use the cumulative distribution function for binomial distribution, \(F(9)\).

\[P(T < 10) = P(T ≤ 9) = F(9)\]

which in R is given by the function pbinom:

```{r}
(score <- pbinom(9, size=20, prob=0.2))
```

So, random guesses will lead to failure in 99.7\% of exams, and the probability of passing is therefore:

\[P(T ≥ 10) = 1−P(T ≤ 9) = 1−F(9) = 1−0.9974 = 0.0026\]

\(~\)

### Plotting the PMF

```{r}
require(ggplot2)

df.p <- data.frame(
  x=1:20, 
  prob=dbinom(1:20, 20, prob=0.2)
  )

plt.pdf <-  
  ggplot(
    data = df.p, 
    aes(x = x, y = prob)) + 
  geom_line() + 
  # geom_ribbon(
  #   data = subset(df, x >= 10 & x <= 20), 
  #   aes(ymax = prob), 
  #   ymin=0, 
  #   fill = "red", 
  #   colour = NA, 
  #   alpha = 0.5
  #   ) + 
  theme_bw() + 
    labs(
         title="Binomial distribution:",              # subtitle="whatever", caption = "whatever", 
         subtitle="Probability mass function",       #, caption = "whatever",
         y="Probability",
         x="Score" 
         ) +
    scale_x_continuous(expand = c(0, 0)) +         # expand = c(0, 0) removes default
    scale_y_continuous(expand = c(0, 0), limits = c(0, 0.23)) +          # gap between axis and (0, 0)  # 
  theme(
      panel.grid = element_line(colour = "white")
    )

plt.pdf
```

\(~\)

### Plotting the CDF

```{r}
df.c <- data.frame(
  x=1:20, 
  prob=pbinom(1:20, 20, prob=0.2)
  )


#x = x, y = prob
#ggplot(df, aes(height)) + stat_ecdf(geom = "point")
#ggplot(df, aes(height)) + stat_ecdf(geom = "step")

plt.cdf <-  
  ggplot(
    data = df.c, 
    aes(x = x, y = prob)) + 
  geom_line() + 
  geom_ribbon(
    data = subset(df.c, x >= 10 & x <= 20),
    aes(ymax = prob),
    ymin=0,
    fill = "darkolivegreen4",
    alpha = 0.25
    ) +
  theme_bw() + 
    labs(
         title="Binomial distribution:", 
         subtitle="Cumulative distribution function",       #, caption = "whatever",
         y="Cumulative probability",
         x="Score" 
         ) +
    scale_x_continuous(expand = c(0, 0)) +         # expand = c(0, 0) removes default
    scale_y_continuous(expand = c(0, 0), limits = c(0, 1.10)) +          # gap between axis and (0, 0)  # 
  theme(
      panel.grid = element_line(colour = "white")
    )

plt.cdf
```



